A recipe for extremely reproducible enrichment analysis

### 2023-03-31 manuscript.Rmd
Update manuscript.Rmd
refining manuscript

To address this, many articles have been written to endorse the use of "virtual machines" or
"Docker containers" to encapsulate the operating system and dependancies along with the research code
to facilitate a high degree of reproducibility [4-16](Perkel 2020; Stein 2010; Dudley & Butte 2010;
Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014; Lewis et al, 2016;
da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018; Hung et al, 2019;
Brito et al, 2020; Piccolo et al, 2021).
Despite the availability of such tools for over a decade, bioinformatics analyses using fully
reproducible virtual machines or containers remains only a small fraction.

![Figure XX: The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")

In this article, we will focus on enrichment analysis, which is one of the most used methods in
computational biology, and involves the summarisation of omics data to reflect biological changes,
such as the detection of pathway activation in development or disease
[17,18](Slonim 2002; Khatri et al, 2012).
While enrichment analysis is undoubtedly a key method for the interpretation of large
data sets, there are concerns that inadvertant misuse of these methods leads to
unreliable results.
These potential problems include (i) the lack of correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[19-22](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the reliability of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script based methods for this type
of work [21](Wijesooriya et al, 2022).
Web-based tools are a potential concern, firstly because sometimes methods are not completely described,
and secondly, algorithms and functional annotation sets undergo regular updates.
These historical versions are typically not archived, which hampers later reproducibility.
In some cases the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay [23](Hennessey & Ge 2013).

@@

But is this extra work really necessary?
Although our previous work suggests that enrichment analyses on the whole lack reproducibility
[5](Wijesooriya et al, 2022), we lack the empirical data confirming this directly.
Therefore before recommending or prescribing Docker-based bioinformatics workflows, it is necessary
to assess directly whether current standards are appropriate, by trying to reproduce the enrichment
analyses of previously published studies, and examining whether the conclusions of those articles
are supported or not.
In the following section we investigate the computational reproducibility of enrichment analyses
shown in a sample of 20 journal articles published in 2019.
This should provide a sound basis for recommendations.

## Methods

### A systematic assessment of reproducibility in enrichment analysis

<Anusuiya to populate>
<references to start at [101](Author Year)>

### A container based approach to addressing reproducibility of enrichment analysis

@@

@@

>>>>

To address this, many articles have been written to endorse the use of "virtual machines" or
"containerisation" to encapsulate the "environment" including the operating system and dependancies
along with the research code to facilitate a high degree of reproducibility [4-16](Perkel 2020;
Stein 2010; Dudley & Butte 2010; Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014;
Lewis et al, 2016; da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018;
Hung et al, 2019; Brito et al, 2020; Piccolo et al, 2021).
Containers also allow for an easy installation process, as all working dependencies are shipped together
and can be installed with a single command.

In addition to shared code, data and containerisation, there are a few ways researchers can make their
work more reproducible.
Firstly, thorough documentation makes the reproducibility process a great deal easier and more accessible
to novice bioinformaticians.
Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
free text explanations together with the code and outputs such as charts and tables.
Secondly, code and data can be linked. This means that the code knows the location of the the data,
avoiding any "file not found" errors.
Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
for each figure sequentially, making the entire manuscript reproducible by executing just one command.
Taken together, computational reproducibility is a spectrum (**Figure 1**).
Currently, bioinformatics work at the gold standard is rare, despite the availability of such tools for
over a decade.

![Figure 1: The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")

In this article, we focus on enrichment analysis, which is one of the most used methods in
computational biology, and involves the summarisation of omics data to reflect biological changes,
such as the detection of pathway activation in development or disease
[17,18](Slonim 2002; Khatri et al, 2012).
While enrichment analysis is undoubtedly a key method for the interpretation of large
data sets, there are concerns that inadvertant misuse of these methods leads to
unreliable results.
These potential problems include (i) the lack of correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[19-22](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the reliability of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script based methods for this type
of work [21](Wijesooriya et al, 2022).
Web-based tools are a potential concern, firstly because sometimes methods, options and parameters
are not completely described, and secondly, algorithms and functional annotation sets undergo
regular updates.
These historical versions are typically not archived, which hampers later reproducibility.
In some cases the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay [23](Hennessey & Ge 2013).

@@

But is this extra work really necessary?
Although our previous work suggests that enrichment analyses generally lack sufficient methodological
detail [5](Wijesooriya et al, 2022), we don't have empirical evidence confirming irreproducibility.
Therefore before recommending or prescribing Docker-based bioinformatics workflows, it is necessary
to assess directly whether current practices are appropriate, by trying to reproduce the enrichment
analyses of previously published studies, and examining whether the conclusions of those articles
are supported or not.
In the following section we investigate the computational reproducibility of enrichment analyses
shown in a sample of 20 journal articles published in 2019.
This should provide a sound basis for recommendations.

## A systematic assessment of reproducibility in enrichment analysis

### Methods

<Anusuiya to populate>
<references to start at [101](Author Year)>
<Wanted to check how this looks in the file>

From a list of 186 PubMed central articles using enrichment analysis that we examined previously
[5](Wijesooriya et al, 2022), we selected 20 articles that fulfilled the following criteria; (1) the
organism under study is human; (2) the tool used was DAVID (Database for Annotation, Visualization,
and Integrated Discovery)[101](Huang et al, 2007); (3) the omics type was gene expression (array or
RNA-seq); and (4) that the gene list was provided in the supplementary data.

The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
special attention to preventing gene name errors from occurring[102](Zeeberg et al, 2004).
The reproducibility analysis was performed using the respective methodologies stated in the methods
sections of the 20 articles with DAVID version 6.8 (March-May 2022).

Briefly, we used the DAVID "Analysis Wizard" and pasted in the provided gene list into the text box,
selecting this as "gene list", together with the appropriate gene identifier type.
Typically this was the official gene symbol.
The species was specified as *Homo sapiens*.
Next we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results”
page.
We followed the steps indicated in the respective papers to select appropriate gene sets investigated
in the articles such as gene ontology (biological process, molecular function), KEGG pathways, etc.
Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
enrichment results, which were saved for future reference.

Firstly, the data presented in the paper was compared with the results we obtained, recording any
discrepancies.
Secondly, the statements made in the abstract, results, and discussion sections were carefully examined
and cross-checked with our results to understand whether those assertions survive reproduction or not.

Each article was critically evaluated and given a score of one (1) to three (3).
Articles rated 1 showed overall poor reproducibility, for example most GO and/or KEGG terms did not
match upon reproduction, and the conclusions of the study were overall not supported.
Aticles were rated 2 when at least 50% of the terms are observed as significant after replication,
according to the original study's significance threshold, and most conclusions were supported.
Articles were rated 3 when they showed a high level of reproducibility, with 80% or more of the terms
shown in the article observed again in reproduction, and all conclusions were supported in reproduction.
Results are summarised in table and bar chart form.

### Results and discussion

* bar graph

* table

*

It is important to note that the DAVID 6.8 server has been replaced by DAVID 2022 server, starting
from June 1, 2022.
For our analysis, DAVID version 6.8 was used for replication procedures.
It was ensured that our analyses got completed by 31 May 2022.
It is no longer possible to reproduce analyses conducted in DAVID 6.8

## A container based approach to addressing reproducibility of enrichment analysis

### Implementation

@@

### Findings

@@

101.  Huang DW, Sherman BT, Tan Q, Collins JR, Alvord WG, Roayaei J, et al. The DAVID Gene Functional Classification Tool: a novel biological module-centric algorithm to functionally analyze large gene lists. Genome Biol [Internet]. 2007;8(9):R183. Available from: http://dx.doi.org/10.1186/gb-2007-8-9-r183
102.  Zeeberg BR, Riss J, Kane DW, Bussey KJ, Uchio E, Linehan WM, et al. Mistaken identifiers: gene name errors can be introduced inadvertently when using Excel in bioinformatics. BMC Bioinformatics [Internet]. 2004;5:80. Available from: http://dx.doi.org/10.1186/1471-2105-5-80
